---
title: "Preparing data from the Global Burden of Disease for multi-state data frame"
author: -" Zapata-Diomedi, belen  <belen.zapata-diomedi@rmit.edu.au>"
        - "Jackson, Chris <chris.jackson@mrc-bsu.cam.ac.uk>"
date:   "`r Sys.time()`"
output:
    html_document:
        code_folding: show
params:
    date:   !r Sys.Date() - 1
---


# Read data 
Data for generation of disbayes inputs and data frame for mslt 
For now, we add outputs from Chris to mslt data frame

BELEN: try to reduce computation burden from functions


```{r,eval=FALSE}
library(tidyverse)
library(tidyr)

### Belen: how to use ~ to make it reproducible??
# relative_path_gbd <- "~/scratch/chronic"
# relative_path_execute <- "~/scratch/chronic/mh-execute"
relative_path_execute <- 'C:/Metahit/mh-execute'
relative_path_gbd <- 'C:/Metahit/mh-mslt/data/gbd/GBD2017/METAHIT/'

```

Read look-up table mapping between local authorities and city regions, and add names for regions that are not city regions

```{r,eval=FALSE}
names_non_cr <- c("United Kingdom", "England", "East Midlands", "East of England", "Greater London", "North East England", 
                  'North West England', "South East England", "South West England", "West Midlands", "Yorkshire and the Humber", 
                  "Northern Ireland", "Scotland", "Wales")
local_government_areas <- read_csv(file.path(relative_path_execute, "inputs/mh_regions_lad_lookup.csv"))
for (i in names_non_cr){
  local_government_areas <- rbind(local_government_areas, rep(i, ncol(local_government_areas)))
}
local_government_areas  <- local_government_areas %>%
    dplyr::filter(!is.na(cityregion)) %>%
    dplyr::rename(location = lad11nm) %>%
    dplyr::mutate(location = gsub('St. Helens', 'St Helens', location))
```

```{r,eval=FALSE}
gbd1 <- read_csv(file.path(relative_path_gbd, "IHME-GBD_2017_DATA-3e0b192d-1.csv"))
gbd2 <- read_csv(file.path(relative_path_gbd, "IHME-GBD_2017_DATA-3e0b192d-2.csv"))
gbd3 <- read_csv(file.path(relative_path_gbd, "IHME-GBD_2017_DATA-3e0b192d-3.csv"))
gbd4 <- read_csv(file.path(relative_path_gbd, "IHME-GBD_2017_DATA-3e0b192d-4.csv"))
gbd5 <- read_csv(file.path(relative_path_gbd, "IHME-GBD_2017_DATA-3e0b192d-5.csv"))
gbd <- rbind(gbd1, gbd2, gbd3, gbd4, gbd5) %>%
    dplyr::filter(location_name %in% local_government_areas$location)
rm(gbd1, gbd2, gbd3, gbd4, gbd5)

```

Read look-up table mapping between local authorities and city regions, and add names for regions that are not city regions

```{r,eval=FALSE}
names_non_cr <- c("United Kingdom", "England", "East Midlands", "East of England", "Greater London", "North East England", 
                  'North West England', "South East England", "South West England", "West Midlands", "Yorkshire and the Humber", 
                  "Northern Ireland", "Scotland", "Wales")
local_government_areas <- read_csv(file.path(relative_path_execute, "inputs/mh_regions_lad_lookup.csv"))
for (i in names_non_cr){
  local_government_areas <- rbind(local_government_areas, rep(i, ncol(local_government_areas)))
}
local_government_areas  <- local_government_areas %>%
    dplyr::filter(!is.na(cityregion)) %>%
    dplyr::rename(location = lad11nm) %>%
    dplyr::mutate(location = gsub('St. Helens', 'St Helens', location))

```
Read disease coding table

```{r,eval=FALSE}
disease_names_execute <- read_csv(file.path(relative_path_execute, "inputs/dose_response/disease_outcomes_lookup.csv")) %>%
    select(GBD_name, acronym) %>%
    mutate(disease = tolower(GBD_name))
    
DISEASE_SHORT_NAMES <- data.frame(disease = tolower(as.character(unique(gbd$cause_name))), 
                                  sname = tolower(abbreviate(unique(gbd$cause_name, max = 2))),
                                  stringsAsFactors = F) %>%
    dplyr::mutate(is_not_dis = ifelse((str_detect(disease, "injuries") |
                                str_detect(disease, "All causes") |
                                str_detect(disease, "Lower respiratory infections")), 
                               1, 0) ) %>%
    dplyr::mutate(is_not_dis = case_when(sname == "allc"  ~  2,
                                  sname == "lwri"  ~  1,
                                  ## Code for major depressive disorder (no deaths) and hypertensive heart disease (no incidence)
                                  sname == "hyhd"  ~  3,
                                  sname == "mjdd"  ~  3,
                                  TRUE  ~  is_not_dis)) %>%
    left_join(disease_names_execute, by="disease") %>%
    dplyr::mutate(acronym = ifelse(str_detect(disease, "injuries"), disease, acronym),
           acronym = word(acronym, 1),
           males = ifelse(disease %in% c("uterine cancer", "breast cancer"), 0, 1),
           females = ifelse(disease %in% "prostate cancer", 0, 1),
           sname = gsub("'", '', sname),
           acronym = ifelse(is.na(acronym), "no_pif", acronym))

saveRDS(DISEASE_SHORT_NAMES, paste0(relative_path_gbd, "DISEASE_SHORT_NAMES.rds"))
```    

Join geographical data to GBD data 

```{r,eval=FALSE}
names(gbd) <- gsub(pattern = "_name", replacement = "", x = names(gbd))
gbd <- gbd %>%
    dplyr::select(-contains("id")) %>%
    dplyr::mutate(cause = tolower(cause)) %>%
    left_join(local_government_areas, by="location")
```

## Calculate population numbers

```{r,eval=FALSE}
gbdnum <- gbd %>%
  dplyr::filter(metric=="Number") %>%
  dplyr::select(measure, location, sex, age, cause, cityregion, Number=val)

gbdval<- gbd %>%
  dplyr::filter(metric=="Rate") %>%
  dplyr::select(measure, location, sex, age, cause, cityregion, val) %>%
  dplyr::mutate(Rate = val / 100000)
gbdpop <- gbdval %>%
   left_join(gbdnum, by=c("measure","location","sex","age","cause", "cityregion")) %>%
   dplyr::mutate(pop = Number / Rate)

```


## Aggregate by city regions (includes regions and countries)

```{r,eval=FALSE}
options(dplyr.summarise.inform = FALSE) 
gbd_agg <-  dplyr::select(gbdpop, measure, sex, age, cause, cityregion, Number, pop) %>%
  group_by(measure, sex, age, cause, cityregion) %>%
  summarise(number = sum(Number), pop = sum(pop)) %>%
  ungroup
```

Calculate rates per one. This step is needed as we add up localities within city regions.

```{r,eval=FALSE}

gbd_agg <-  gbd_agg %>%
 dplyr::mutate(rate = number / pop)
```

Create age variables: from, to and mid age interval
```{r,eval=FALSE}
gbd_agg <- gbd_agg %>%
  extract(age, c("from_age", "to_age"), "(.+) to (.+)", remove=FALSE, convert=TRUE) %>%
  dplyr::mutate(from_age = case_when(age =="95 plus"  ~  95L,
                              age =="Under 5"  ~  0L,
                              TRUE  ~  from_age),
         to_age = case_when(age == "95 plus"  ~  99L,
                            age == "Under 5"  ~  4L,
                            TRUE  ~  to_age)) %>%
  dplyr::mutate(age_cat = from_age + 2)
```

## Generate data frames for MSLT, excluding incidence and case fatality

Create wider data frame for measure and cause combinations
```{r,eval=FALSE}

gbd_agg_wider <- gbd_agg %>% 
  dplyr::mutate(disease = tolower(abbreviate(cause))) %>%
  dplyr::mutate(age_sex = paste0(age, sex)) %>%
  group_by(age_sex) %>% 
  pivot_wider(id_cols = c(measure, sex, age, disease, cityregion, number, pop, rate, age_cat, age_sex), 
             names_from = c(measure, disease), values_from = c(rate, number))
```

Add names to columns

``` {r,eval=FALSE}
gbd <- gbdpyrcr %>%
  filter(measure %in% c("Deaths","Incidence", "Prevalence")) %>%
  mutate(measure = fct_recode(measure, mort="Deaths", inc="Incidence", prev="Prevalence")) %>% 
  pivot_wider(names_from=measure, values_from=c("num","denom"))
```